---
title: Point Shape Options in ggplot
author: ''
date: '2019-03-24'
slug: point-shape-options-in-ggplot
categories:
  - coding
tags:
  - R
  - ggplot
  - geom_point
showDate: true
draft: false
---

I'm familiar enough with ggplot that I can make a quick plot pretty easily in most cases.^[Yes, I spoke too soon. The plots on this post took way longer to make than anticipated.] But when it comes to the various plot aesthetics, like adjusting the legend position or rotating axis tick labels, I always have to look them up. Today, I will be writing about one of these pesky things: looking up the point shape options for `geom_point`. I feel like the available documentation for this isn't great, so I'm writing my own reference.

# How to Change the Shape

By default, `shape = 19` (a filled circle). You can change the number to plot different shapes, i.e. `geom_point(shape = x)`. If you want to change point shapes based on a grouping variable, then first set the shape with the grouping variable in `geom_point` and then use `scale_shape_manual` to choose the desired shapes (optional). Below is an example.

```{r message = F, warning = F}
library(ggplot2)

x = 1:5
y = rnorm(5)
group = as.factor(c(1, 1, 2, 2, 2))
dt = data.frame(x, y, group)

ggplot(dt, aes(x, y)) +
  geom_point(aes(shape = group, color = group), size = 3) + 
  scale_shape_manual(values = c(15, 16)) + 
  theme_minimal()
```

The first 26 options available for point shapes, i.e. shapes 0 to 25, are the ones you are most likely to use.^[There are additional options available, i.e. shapes 32 to 127, which correspond to the ASCII characters. I think it's quite unlikely I'll ever use them, but if you're interested, you can view them [here](http://sape.inf.usi.ch/quick-reference/ggplot2/shape).] I have created two keys below for the different shape options. The first is ordered by the shape number and the second is ordered by the geometric shape.

# Shape Options (Ordered By Number)

```{r echo = F}
y = rep(5:1, times = c(5, 5, 5, 6, 5))
x = c(1:5, 1:5, 1:5, 1:6, 1:5)
shape = 0:25
dt = data.frame(x, y, shape)
ggplot(dt, aes(x, y, label = shape)) +
  geom_point(aes(shape = as.factor(shape)), size = 5,
             color = "#ec2027", fill = "#fbd2d3") + 
  geom_text(nudge_y = 0.2)  + 
  annotate("text", x = 0, y = 5, label = "Outline") +
  annotate("text", x = 0, y = 2, label = "Fill") +
  annotate("text", x = 0, y = 1, label = "Both") +
  scale_shape_manual(values = dt$shape) + 
  guides(shape = F) +
  theme_void()
```

Notes: 

* Shapes 0 to 14 are outline only: use `color` to change colors
* Shapes 15 to 20 are fill only: use `color` to change colors
* Shapes 21 to 25 are outline + fill: use `color` to change the outline color and `fill` to change the fill color

# Shape Options (Ordered by Shape)

```{r echo = F}
y = rep(5:1, times = c(7, 5, 4, 6, 4))
x = c(1:7, 1:5, 1:4, 1:6, 1:4) 
shape = c(1, 10, 13, 16, 19, 20, 21,
          2, 6, 17, 24, 25,
          5, 9, 18, 23,
          0, 7, 12, 14, 15, 22,
          3, 4, 8, 11)
shape_factor = as.factor(0:25)
dt = data.frame(x, y, shape, shape_factor)
ggplot(dt, aes(x, y, label = shape)) +
  geom_point(aes(shape = shape_factor), size = 5,
             color = "#ec2027", fill = "#fbd2d3") + 
  geom_text(nudge_y = 0.2) + 
  annotate("text", x = 0, y = 5, label = "Circles") +
  annotate("text", x = 0, y = 4, label = "Triangles") +
  annotate("text", x = 0, y = 3, label = "Diamonds") +
  annotate("text", x = 0, y = 2, label = "Squares") +
  annotate("text", x = 0, y = 1, label = "Other") +
  scale_shape_manual(values = dt$shape) + 
  coord_cartesian(xlim = c(-0.5, 8)) +
  guides(shape = F) +
  theme_void()
```

I often find that I'm looking for a particular shape when plotting, so I've also arranged the options by their geometric shape above.
