---
title: Introduction to Regular Expressions (regex) in R
author: ''
date: '2020-01-08'
slug: introduction-to-regular-expressions-regex-in-r
categories:
  - coding
  - R
tags:
  - regex
  - R
showDate: yes
draft: true
---

If you haven't used [regular expressions](https://en.wikipedia.org/wiki/Regular_expression) (regex) before, they are basically a way to write search patterns for strings. I've always found them to be inscrutable and unintuitive, so even though the search pattern I have in mind is usually very simple and should, in theory, require only a basic regex, I always have to Google what the correct syntax is. Today, I'm going to try to solve this problem by writing my own guide to regex.^[As I continue to familiarize myself with regex, I will update this post.]

## R commands

First, let's go over the R functions that you are likely going to use regex with:

* `grep`: Search for matches of the regex in a vector and returns the indices of elements that match if `value = F` or the elements themselves if `value = T`
* `grepl`: Search for matches of the regex in a vector and return a boolean vector
* `sub`: Replace the first match of the regex with a new string
* `gsub`: Replace all matches of the regex with a new string
* `regexpr`: Return starting position (-1 if none) and length of first match
* `grepexpr`: Return starting positions (-1 if none) and lengths of all matches
* `regmatches`: Extract matched substrings obtained by `regexpr`. If `invert = T`, extract the substrings that don't match to the regex.


## Examples

To demonstrate how the different regular expressions work, I will use the following example vector.

```{r echo = F, message = F}
library(dplyr)
```

```{r}
example_vec = c("The big ocean", "The big Ocean", "The big ocean.", "The big. ocean!", "The big.. ocean!")
example_vec %>% cat(., sep = "\n")
```

### Anchors

* `^`: Start of string
* `$`: End of string

```{r eval = T}
# Find elements that end with "ocean"
grep("ocean$", example_vec, value = T) %>% cat(., sep = "\n")
```

### More Characters

* `.`: Match any character
* `[a-z]`, `[0-9]`, `[Tt]`, `[[:punct:]]`, `[[:alpha:]]`, `[[:lower:]]`, `[[:upper:]]`,`[[:digit:]]`: Search for any of the options in the brackets (i.e. "character classes")
* `\\`: Escape a character.^[For more details about why a double backslash is necessary,  see https://stackoverflow.com/questions/27721008/how-do-i-deal-with-special-characters-like-in-my-regex.] For example, to search for a period, you need to use `\\.`.
* `|`: "OR" operand
* `^`: "NOT" operand

```{r eval = T}
# Find elements that end with "ocean" plus any character
grep("ocean.$", example_vec, value = T) %>% cat(., sep = "\n")
# Find elements that end with "ocean" or "Ocean"
grep("[Oo]cean$", example_vec, value = T) %>% cat(., sep = "\n")
# Find elements that end with a punctuation
grep("[[:punct:]]$", example_vec, value = T) %>% cat(., sep = "\n")
# Find elements that end with a period
grep("\\.$", example_vec, value = T) %>% cat(., sep = "\n")
# Find elements that end with "ocean" or "Ocean"
grep("(ocean$|Ocean$)", example_vec, value = T) %>% cat(., sep = "\n")
# Find elements that end with "cean" but not "Ocean"
grep("[^O]cean$", example_vec, value = T) %>% cat(., sep = "\n")
```

### Repetition

* `*`: Repeat 0 or more times
* `+`: Repeat 1 or more times
* `{2}`: Repeat exactly 2 times
* `{2, 4}`: Repeat 2 to 4 times

```{r eval = T}
# Find elements that start with "The big."
grep("^The big\\.+", example_vec, value = T) %>% cat(., sep = "\n")
# Replace substring beginning with the first period, if it exists
gsub("\\.(.*)$", "REPLACEMENT", example_vec) %>% cat(., sep = "\n")
# Replace substring beginning with the last period, if it exists
gsub("\\.[^\\.]*$", "REPLACEMENT", example_vec) %>% cat(., sep = "\n")
# Find matched substrings beginning with the first period
regmatches(example_vec, regexpr("\\.(.*)$", example_vec)) %>% cat(., sep = "\n")
# Find the substrings that don't match to "big.."
regmatches(example_vec, regexpr("big(\\.{2})", example_vec), invert = T)
```

## Resources

Here are some resources and references I used while writing this blog post.

* You can test or look up your regex at https://regex101.com/#python
* [RStudio cheat sheet](https://rstudio.com/wp-content/uploads/2016/09/RegExCheatsheet.pdf)
* [JHU Data Science lecture on regex](https://jhu-advdatasci.github.io/2018/lectures/10-regex.html)
