---
title: Introduction to Regular Expressions (regex) in R
author: ''
date: '2020-01-08'
slug: introduction-to-regular-expressions-regex-in-r
categories:
  - coding
  - R
tags:
  - regex
  - R
showDate: no
draft: true
---

[Regular expressions](https://en.wikipedia.org/wiki/Regular_expression) (regex) are search patterns. I've always found them to be inscrutable and unintuitive, so even though the search pattern I have in mind is usually very simple and should, in theory, require only a basic regex, I always have to Google what the correct syntax is. Today, I'm going to try to solve that problem by writing a guide to regex in R.

## R commands

First, let's go over the R functions that you are likely to use regex for:

* `grep`: Search for matches of the regex in a vector and returns the indices of elements that match if `value = F` or the elements themselves if `value = T`
* `grepl`: Search for matches of the regex in a vector and returns a boolean vector
* `sub`: Replaces the first match of the regex with a new string
* `gsub`: Replaces all matches of the regex with a new string


## Examples

To demonstrate how the different regular expressions work, I will use the following example vector.

```{r}
example_vec = c("The big ocean", "The big Ocean", "The big ocean.", "The big. ocean!", "The big.. ocean!")
print(example_vec)
```

### Anchors

* `^`: Start of string
* `$`: End of string

```{r}
# Find elements that end with "ocean"
grep("ocean$", example_vec, value = T)
```

### More Characters

* `.`: Match any character with a period
* `[a-z]`, `[0-9]`, `[Tt]`, `[[:punct:]]`, `[[:alpha:]]`, `[[:lower:]]`, `[[:upper:]]`,`[[:digit:]]`: Search for any of the options in the brackets (i.e. "character classes")
* `\\`: Escape a character.^[For more details about why a double backslash is necessary, https://stackoverflow.com/questions/27721008/how-do-i-deal-with-special-characters-like-in-my-regex.] For example, to search for a period, you need to use `\\.`.
* `|`: "OR" operand
* `^`: "NOT" operand

```{r}
# Find elements that end with "ocean" or "Ocean"
grep("[Oo]cean$", example_vec, value = T)
# Find elements that end with a punctuation
grep("[[:punct:]]$", example_vec, value = T)
# Find elements that end with a period
grep("\\.$", example_vec, value = T)
# Find elements that end with "ocean" or "Ocean"
grep("(ocean$|Ocean$)", example_vec, value = T)
# Find elements that end with "cean" but not "Ocean"
grep("[^O]cean$", example_vec, value = T)
```

### Repetition

* `*`: Repeat 0 or more times
* `+`: Repeat 1 or more times
* `{3}`: Repeat exactly 3 times
* `{2, 4}`: Repeat 2 to 4 times

```{r}
# Find elements that start with "The big"
grep("^The big\\.*", example_vec, value = T)
# Find elements that start with "The big."
grep("^The big\\.+", example_vec, value = T)
# Remove substring beginning with the first period, if it exists
gsub("\\.(.*)$", "", example_vec)
# Remove substring beginning with the last period, if it exists
gsub("\\.[^\\.]*$", "", example_vec)
```

